Variance reduction for MCMC methods via martingale representations

### 2020-03-11 aap/Revision/vr-mcmc-revision-new.tex
update

\begin{proof}
Applying~\eqref{eq:06062018a1} recursively, we get
$$
x_p\le\ol C_0\prod_{k=1}^p \alpha_k
+\gamma\sum_{r=1}^p \epsilon_r
\prod_{k=r+1}^p \alpha_k,
$$
where we use the convention $\prod_{k=p+1}^p:=1$.
Substituting estimate~\eqref{eq:06062018a2}
into the right-hand side, we obtain
$$
x_p\le\left(\ol C_0+\ol C_1
\gamma\sum_{r=1}^p  \prod_{k=1}^r \alpha_k
\right)
\prod_{k=1}^p \alpha_k,
$$
which, together with~\eqref{eq:06062018a3}, completes the proof.
\end{proof}
In what follows, we use the notation
\begin{equation}
\label{eq:definition-alpha}
\alpha_k=1-\gamma \mu'(X_{k-1}),\quad k\in\mathbb N.
\end{equation}
The assumption \eqref{eq:smooth-mu} implies that  $|\mu'(x)|\leq B_\mu$ for some constant $B_\mu>0$ and all $x\in \mathbb{R}^d$. Without loss of generality we suppose that $\gamma B_\mu<1.$
\begin{lem}\label{lem:06062018a2}
Under assumptions of Theorem~\ref{th:mr},
for all natural $r\le K$ and $l\le p$, there exist constants $C_r$ (not depending on $l$ and $p$) such that
\begin{equation}
\label{eq:08062018b2}
\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p (1-\gamma\mu'(X_{k-1}))\quad \text{a.s.}
\end{equation}
where $\partial_{y_l}^r X_p$ is defined in \eqref{eq:definition-differential-f-p}. Moreover, we can choose $C_1=1$.
\end{lem}
\begin{lem}\label{lem:06062018a3}
Under assumptions of Theorem~\ref{th:mr},
for all natural $r\le K$, $j\ge l$ and $p>j$, we have
\begin{equation}\label{eq:08062018b3}
\left|\partial_{y_j} \partial_{y_l}^r X_p\right|
\le c_r\prod_{k=l+1}^p (1-\gamma\mu'(X_{k-1})), \quad \text{a.s.}
\end{equation}
with some constants $c_r$
not depending on $j$, $l$ and $p$,
while, for $p\le j$, it holds
$\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}=0$.
\end{lem}

@@

\partial_{y_{l}}X_{p}
=\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{l}}X_{p-1}

@@

\partial_{y_{j}}\partial_{y_{l}}X_{p}
=\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}X_{p-1}-\gamma\mu''(X_{p-1})\partial_{y_{j}}X_{p-1}\partial_{y_{l}}X_{p-1}.

@@

|\partial_{y_{j}}\partial_{y_{l}}X_{p}|
&\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X_{p-1}|

@@

&\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X_{p-1}|

@@

to bound $|\partial_{y_{j}}\partial_{y_{l}}X_{p}|$

@@

(notice that $\partial_{y_j}\partial_{y_l}X_j=0$, that is,

@@

\left|\partial_{y_{j}}\partial_{y_{l}}^{k}X_{p}\right|

@@

\partial_{y_{l}}^{r}X_{p}
&=\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{l}}^{r}X_{p-1}

@@

&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X_{p-1})\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}},

@@

$\partial_{y_l}^r X_{p-1}$

@@

&\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}
=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}-\gamma\mu''(X_{p-1})\partial_{y_{l}}^{r}X_{p-1}\partial_{y_{j}}X_{p-1}

@@

&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1}+1)}(X_{p-1})\partial_{y_{j}}X_{p-1}\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}

@@

&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X_{p-1})\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}\right]

@@

&\hspace{1em} =\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}+\gamma\epsilon_{l,j,p},

@@

\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}\right]&=&\sum_{s=1}^{r-1}\frac{m_{s}}{s!}\left(\frac{\partial_{y_{l}}^{s}X_{p-1}}{s!}\right)^{m_{s}-1}\partial_{y_{j}}\partial_{y_{l}}^{s}X_{p-1}

@@

&& \times \prod_{k\le r-1,\,k\neq s}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}.

@@

|\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}|
\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}|

@@

to bound $|\partial_{y_{j}}\partial_{y_{l}}^r X_{p}|$

@@

(notice that $\partial_{y_j}\partial_{y_l}^r X_j=0$, that is,

>>>>

\begin{proof}
Applying~\eqref{eq:06062018a1} recursively, we get
$$
x_p\le\ol C_0\prod_{k=1}^p \alpha_k
+\gamma\sum_{r=1}^p \epsilon_r
\prod_{k=r+1}^p \alpha_k,
$$
where we use the convention $\prod_{k=p+1}^p:=1$.
Substituting estimate~\eqref{eq:06062018a2}
into the right-hand side, we obtain
$$
x_p\le\left(\ol C_0+\ol C_1
\gamma\sum_{r=1}^p  \prod_{k=1}^r \alpha_k
\right)
\prod_{k=1}^p \alpha_k,
$$
which, together with~\eqref{eq:06062018a3}, completes the proof.
\end{proof}
In what follows, we use the notation
\begin{equation}
\label{eq:definition-alpha}
\alpha_k=1-\gamma \mu'(X_{k-1}^x),\quad k\in\mathbb N.
\end{equation}
The assumption \eqref{eq:smooth-mu} implies that  $|\mu'(x)|\leq B_\mu$ for some constant $B_\mu>0$ and all $x\in \mathbb{R}^d$. Without loss of generality we suppose that $\gamma B_\mu<1.$
\begin{lem}\label{lem:06062018a2}
Under assumptions of Theorem~\ref{th:mr},
for all natural $r\le K$ and $l\le p$, there exist constants $C_r$ (not depending on $l$ and $p$) such that
\begin{equation}
\label{eq:08062018b2}
\left|\partial_{y_l}^r X^x_p \right| \le C_r\prod_{k=l+1}^p (1-\gamma\mu'(X^x_{k-1}))\quad \text{a.s.}
\end{equation}
where $\partial_{y_l}^r X^x_p$ is defined in \eqref{eq:definition-differential-f-p}. Moreover, we can choose $C_1=1$.
\end{lem}
\begin{lem}\label{lem:06062018a3}
Under assumptions of Theorem~\ref{th:mr},
for all natural $r\le K$, $j\ge l$ and $p>j$, we have
\begin{equation}\label{eq:08062018b3}
\left|\partial_{y_j} \partial_{y_l}^r X^x_p\right|
\le c_r\prod_{k=l+1}^p (1-\gamma\mu'(X^x_{k-1})), \quad \text{a.s.}
\end{equation}
with some constants $c_r$
not depending on $j$, $l$ and $p$,
while, for $p\le j$, it holds
$\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p}=0$.
\end{lem}

@@

\partial_{y_{l}}X^x_{p}
=\left[1-\gamma\mu'(X^x_{p-1})\right]\partial_{y_{l}}X^x_{p-1}

@@

\partial_{y_{j}}\partial_{y_{l}}X^x_{p}
=\left[1-\gamma\mu'(X^x_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}X^x_{p-1}-\gamma\mu''(X^x_{p-1})\partial_{y_{j}}X^x_{p-1}\partial_{y_{l}}X^x_{p-1}.

@@

|\partial_{y_{j}}\partial_{y_{l}}X^x_{p}|
&\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X^x_{p-1}|

@@

&\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X^x_{p-1}|

@@

to bound $|\partial_{y_{j}}\partial_{y_{l}}X^x_{p}|$

@@

(notice that $\partial_{y_j}\partial_{y_l}X^x_j=0$, that is,

@@

\left|\partial_{y_{j}}\partial_{y_{l}}^{k}X^x_{p}\right|

@@

\partial_{y_{l}}^{r}X^x_{p}
&=\left[1-\gamma\mu'(X^x_{p-1})\right]\partial_{y_{l}}^{r}X^x_{p-1}

@@

&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X^x_{p-1})\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X^x_{p-1}}{k!}\right)^{m_{k}},

@@

$\partial_{y_l}^r X^x_{p-1}$

@@

&\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p}
=\left[1-\gamma_{p}\mu'(X^x_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p-1}-\gamma\mu''(X^x_{p-1})\partial_{y_{l}}^{r}X^x_{p-1}\partial_{y_{j}}X^x_{p-1}

@@

&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1}+1)}(X^x_{p-1})\partial_{y_{j}}X^x_{p-1}\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X^x_{p-1}}{k!}\right)^{m_{k}}

@@

&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X^x_{p-1})\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X^x_{p-1}}{k!}\right)^{m_{k}}\right]

@@

&\hspace{1em} =\left[1-\gamma\mu'(X^x_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p-1}+\gamma\epsilon_{l,j,p},

@@

\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X^x_{p-1}}{k!}\right)^{m_{k}}\right]&=&\sum_{s=1}^{r-1}\frac{m_{s}}{s!}\left(\frac{\partial_{y_{l}}^{s}X^x_{p-1}}{s!}\right)^{m_{s}-1}\partial_{y_{j}}\partial_{y_{l}}^{s}X^x_{p-1}

@@

&& \times \prod_{k\le r-1,\,k\neq s}\left(\frac{\partial_{y_{l}}^{k}X^x_{p-1}}{k!}\right)^{m_{k}}.

@@

|\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p}|
\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p-1}|

@@

to bound $|\partial_{y_{j}}\partial_{y_{l}}^r X^x_{p}|$

@@

(notice that $\partial_{y_j}\partial_{y_l}^r X^x_j=0$, that is,
